---
title: "Backend and provenance"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Backend and provenance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`cudagraphR` keeps two different histories:

- **source provenance** describes how the nearest-neighbour input was
  computed; and
- **current compute provenance** describes graph assembly or community
  detection performed by `cudagraphR`.

These histories must not be confused. Neighbours can originate on CUDA while
the current graph and community algorithms still run on the CPU.

## A runnable CPU graph

The core graph constructor accepts a neighbour list, so this example needs no
GPU and no optional clustering package.

```{r cpu-graph}
library(cudagraphR)

vertex_names <- paste0("cell_", 1:6)
index <- matrix(
  c(
    2, 3,
    1, 3,
    1, 2,
    5, 6,
    4, 6,
    4, 5
  ),
  nrow = 6,
  byrow = TRUE,
  dimnames = list(vertex_names, c("neighbor_1", "neighbor_2"))
)
distance <- matrix(
  rep(c(1, 1.25), 6),
  nrow = 6,
  byrow = TRUE,
  dimnames = dimnames(index)
)
neighbors <- list(
  index = index,
  distance = distance,
  device = "cpu"
)

graph <- cuda_knn_graph(
  neighbors,
  weighting = "gaussian",
  symmetrize = "mutual"
)

graph
graph$source_device
graph$compute_device
cuda_provenance(graph)
```

`graph$source_device` describes the neighbour input. In contrast,
`cuda_provenance(graph)` describes only the current `graph_assembly` stage:
it uses the CPU and the `Matrix` backend. A plain list has no earlier
stage-level provenance, so `graph$source_provenance` is `NULL`.

## Preserve an upstream history

When `cudalearnr` is installed, a `cuda_knn` result supplies its own distance
and neighbour-selection stages. `cuda_knn_graph()` retains those stages
separately from graph assembly.

```{r upstream-provenance}
if (requireNamespace("cudalearnr", quietly = TRUE)) {
  set.seed(1)
  observations <- matrix(rnorm(24 * 4), 24, 4)
  rownames(observations) <- paste0("observation_", seq_len(nrow(observations)))

  neighbors_cpu <- cudalearnr::cuda_knn(
    observations,
    k = 3,
    batch_size = 8,
    device = "cpu"
  )
  graph_from_knn <- cuda_knn_graph(neighbors_cpu)

  graph_from_knn$source_device
  graph_from_knn$source_provenance
  cuda_provenance(graph_from_knn)
}
```

The two provenance objects answer different questions:

- `graph_from_knn$source_provenance` explains the upstream distance and
  neighbour-selection work;
- `cuda_provenance(graph_from_knn)` explains graph assembly performed now.

## Community detection remains on the CPU

Louvain and Leiden use the optional `igraph` package. They preserve the graph
assembly record as source provenance, but their current
`community_detection` stage is always CPU.

```{r communities}
if (requireNamespace("igraph", quietly = TRUE)) {
  communities <- cuda_louvain(graph)

  communities$membership
  communities$source_device
  communities$source_provenance
  cuda_provenance(communities)
}
```

For a `cuda_communities` result:

- `source_device` still identifies the original neighbour source;
- `source_provenance` is the graph-assembly stage;
- `upstream_provenance` retains any provenance supplied by the neighbour
  result; and
- `cuda_provenance(communities)` reports the current CPU `igraph` stage.

Changing the source device does not make graph assembly, Louvain, or Leiden
GPU-native.

## Optional CUDA source and the hardware gate

The following block runs only when both a CUDA-enabled R `torch` runtime and
`cudalearnr` are available. It requires CUDA explicitly rather than using
`device = "auto"`, then verifies that CUDA neighbour computation is followed
by fixed-CPU graph assembly.

The cudaverse NVIDIA workflow sets `CUDAVERSE_REQUIRE_CUDA=true`. Under that
gate, missing CUDA hardware or a missing integration dependency is an error,
not a skipped test.

```{r optional-cuda}
require_cuda <- identical(
  tolower(Sys.getenv("CUDAVERSE_REQUIRE_CUDA", "false")),
  "true"
)
cuda_ready <- cudatensr::cuda_available()
knn_ready <- requireNamespace("cudalearnr", quietly = TRUE)

if (require_cuda && (!cuda_ready || !knn_ready)) {
  stop(
    "The hardware gate requires CUDA and the cudalearnr integration."
  )
}

if (cuda_ready && knn_ready) {
  set.seed(2)
  cuda_input <- matrix(rnorm(32 * 4), 32, 4)
  rownames(cuda_input) <- paste0("cuda_observation_", seq_len(nrow(cuda_input)))

  neighbors_cuda <- cudalearnr::cuda_knn(
    cuda_input,
    k = 3,
    batch_size = 8,
    device = "cuda"
  )
  graph_cuda <- cuda_knn_graph(neighbors_cuda)
  graph_cuda_provenance <- cuda_provenance(graph_cuda)

  stopifnot(
    identical(graph_cuda$source_device, "cuda"),
    identical(graph_cuda$compute_device, "cpu"),
    identical(graph_cuda_provenance$stage, "graph_assembly"),
    identical(graph_cuda_provenance$device, "cpu")
  )

  graph_cuda$source_provenance
  graph_cuda_provenance

  if (requireNamespace("igraph", quietly = TRUE)) {
    communities_cuda_source <- cuda_louvain(graph_cuda)
    stopifnot(
      identical(communities_cuda_source$compute_device, "cpu"),
      identical(
        cuda_provenance(communities_cuda_source)$device,
        "cpu"
      )
    )
  }
}
```

Use `device = "cuda"` when CUDA execution is a requirement. Use
`device = "auto"` only when a recorded CPU fallback is acceptable.
